---
title: "Advanced explanations of the `motifcounter` package"
author: "Wolfgang Kopp"
date: "`r Sys.Date()`"
output:
    prettydoc::html_pretty:
        toc: TRUE
        number_sections: TRUE
        fig_caption: yes
        fig_height: 4
        fig_width: 6
vignette: >
    %\VignetteIndexEntry{Advanced explanations of the `motifcounter` package}
    %\VignetteEngine{knitr::rmarkdown}
    %\VignetteEncoding{UTF-8}
bibliography: literature.bib
---


```{r include=FALSE}
library(knitr)
knit_hooks$set(plot = function(x, options) {
    paste('<figure><img src="',
        opts_knit$get('base.url'), paste(x, collapse = '.'),
        '"><figcaption>', options$fig.cap, '</figcaption></figure>',
        sep = '')
})
library(motifcounter)
library(MotifDb)
library(seqLogo)
opts_chunk$set(fig.path="fig/")
```

This vignette is intended to shed a more detailed light
on the methods used under the hood of `motifcounter`.
In particular, we'll explain the approximative models for the
distribution of the number of motif hits.

### Inspect the distribution of the number of motif hits

Inspecting the distribution of the number of motif hits 
might be of interest in order to assess the effect size of motif enrichment.
For example, how do the quantiles of the distribution
relate to the observed number of motif hits?

In this section, we investigate the compound Poisson 
approximation, the combinatorial model and the empirical distribution 
for a perfectly palindromic motif. 

As usual we start by obtaining a background model
```{r}
order=1
file=system.file("extdata","seq.fasta", package="motifcounter")
seqs=Biostrings::readDNAStringSet(file)
bg=readBackground(seqs,order)
```

```{r fig.show=TRUE, fig.width=4, fig.height=3, fig.cap="Palindromic motif"}
motiffile=system.file("extdata",as.character("x32.tab"),
                    package="motifcounter")

motif=t(as.matrix(read.table(motiffile)))
seqLogo(motif)

```

The distribution of the number of motif hits for a set of 100 sequences
of length 100 bp each are shown in the following

```{r fig.show=TRUE, fig.cap="Distribution of the number of hits"}

regions=rep(100,100)
# generate an empirical distribution using 100 samples of 1kb sequences each
simdist=motifcounter:::simulateNumHitsDist(motif,bg,regions,100,singlestranded=FALSE)


op=motifcounter:::probOverlapHit(motif,bg,singlestranded=FALSE)
cpdist=motifcounter:::compoundPoissonDist(regions,op)
combdist=motifcounter:::combinatorialDist(regions,op)
maxhits=80

plot(0:(maxhits+1), cpdist$dist[1:(maxhits+2)],
        col="blue",ylab="Probability",xlab="Num. of hits")
points(0:(maxhits+1),combdist$dist[1:(maxhits+2)],col="red")
points(0:(maxhits+1),simdist$dist[1:(maxhits+2)],col="gray")
legend(50,0.08,c("comp. Poiss.", "combinatorial", "empirical"), 
        col=c("blue","red","gray"), lty=c(1,1,1))

```

**Note:** Scanning both DNA strands for motif hits (at $\alpha=0.001$) can only
yield an even number of motif hits, as a hit on the forward 
strand is always coupled with a hit on the reverse strand. 
Therefore, odd numbers of motif hits are associated 
with a probability of (almost) zero.
